Can we distinguish modes of selective interactions using linkage disequilibrium?

### 2021-02-20 Manuscript.Rmd
Draft and analyses started!

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
bibliography: "references.bib"
abstract: The abstract (but it could also be a normal section).
title: "Paper title"
author:
  - Author 1^[Institution 1, email1]
  - Author 2^[Institution 2, email2]
---

@@

@@

This paper is neat [@Ragsdale2020-hv].

@Ragsdale2020-hv is still a neat paper if we cite it like this.

```{r figure, echo=F, fig.cap="A figure", fig.align="center", out.width="50%"}
knitr::include_graphics("Model.png")
```

See Figure \@ref(fig:figure).

See Supplementary Figure \@ref(fig:figure).

>>>>

---
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: no
    number_sections: false
    keep_tex: true
    includes:
      in_header: "preamble.tex"
    citation_package: natbib
bibliography: "references.bib"
biblio-style: "genetics"
title: Can we distinguish modes of selective interactions using linkage disequilibrium?
author:
header-includes:
- |
  ```{=latex}
  \usepackage{authblk}
  \author{Aaron P. Ragsdale \\ aaronpeaceragsdale@gmail.com}
  \affil{Department of Human Genetics, McGill University, Montreal, Canada \\
         National Laboratory of Genomics for Biodiversity, Irapuato, Mexico \\
         }
  ```
date: \today
---

@@

\begin{abstract} Selection acting on a mutation interferes and interacts with
evolutionary processes at nearby loci, causing allele frequency and correlation
patterns between pairs of selected mutations to deviate from expected
single-locus dynamics. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over interpretation of
observations. Interpreting this data is hindered by the relative lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection, with arbitrary
epistasis and dominance, and variable recombination and population sizes. I use
this to explore how different models of epistasis and dominance affect expected
signed LD, including for non-steady-state demography relevant to human
populations. Finally, I use whole-genome sequencing data to asses how well we
can differentiate models of selective interference in practice, and find [[that
while xxx, within classes/domains, we see...]]. \end{abstract}

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The average number of generations that a
deleterious mutation segregates in a population depends on the strength of
selection against genomes that carry it, with very damaging mutations kept at
low frequencies in the population and purged relatively rapidly. But in the
time between mutation and fixation or loss, selected variants, both beneficial
and damaging, can dramatically impact patterns of variation in nearby linked
regions. This distortion away from neutral expectations has been well
documented in practically every taxon and population for which we have genomic
data. However, questions remain about the primary mode of interactions between
multiple linked variants and their joint effects on genome-wide patterns of
diversity.

In their foundational paper, @Hill1966-gv recognized that linked negatively
selected variants reciprocally impede the efficacy of population to remove
those mutations. In general, linked selection reduces the fixation probability
of advantageous mutations and increases that of deleterious mutations, compared
to expectations under single-locus dynamics [@Birky1988-jm]. Allele frequency
dynamics of linked selected variants are also predicted to deviate from
expectations without interference. Under a multiplicative fitness model, where
the fitness reduction of a haplotype carrying multiple deleterious variants is
equal to the product of the fitness reduction of each mutation independently,

One (!!) paragraph on epistasis, dominance, etc and overall importance. Thus,
there is a lot of interest in disentangling and distinguishing between these
processes.

## Empirical observations

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are induced within a target gene and fitness, or some protein function, is
experimentally measured [@Romero2009-yi; @Steinberg2016-is]. For example, using
the model system of the TEM-1 $\beta$-lactamase gene in *E. coli*,
@Bershtein2006-bk found evidence for synergistic epistasis, where multiple
deleterious mutations had a greater effect on fitness than the multiplication
of the observed effects of the individual mutations. The scale of mutation
scanning experiments continues to improve dramatically, promising greater
resolution of the fitness landscape in such model systems.

In most natural populations, directed mutation studies are not possible, and we
must turn to population genetic approaches to infer selective interactions
between observed mutations. Motivated by theory that linked negatively selected
mutations will display negative LD due to interference [@Hill1966-gv], and that
epistasis will drive expected LD away from zero, a number of recent studies
have used patterns of LD within classes of putatively selected variants to
infer modes of selective interactions. In a notable study from
@Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster more
than expected along lineages in the *Drosophilid* species complex, and that
those clustered mutations tended to preserve the charge of the protein and were
in positive LD compared to pairs including synonymous mutations, suggesting
that compensatory nonsynonymous variants were more generally tolerated.

More recently, @Sohail2017-zq observed negative LD between loss-of-function
variants in protein-coding genes (such as stop gains and losses, frameshifts,
and other nonsense mutations) in both human and fruit fly populations, from
which they proposed that widespread synergistic epistasis between these
mutations. Within the past year or two, both @Sandler2020-vk and @Garcia2020-zn
have reevaluated patterns of LD between coding variants in humans, fruit flies,
and *Capsella grandiflora*, and suggested interference and dominance may
instead be driving patterns of LD [@Garcia2020-zn] or questioned whether LD
between loss-of-function variants is significantly different from zero
[@Sandler2020-vk].

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function variants,
their rarity and low frequency means that statistical measurement of LD and
other diversity measures are quite noisy. Second, comparisons are based on
theory that are confined to limiting and simplistic assumptions, including
steady-state demography, simple selection and interaction models, or unlinked
loci. To generate predictions under more complex models, we rely on expensive
forward simulations. These can be great for building intuition or testing
inference methods, but do not efficiently provide expectations for quantities
of interest across parameter regimes of interest. Theoretical and numerical
studies of haplotype frequencies and LD under general selective interaction
models would be of great benefit.

## Existing theory and numerical methods

In addition to impacting surrounding neutral variation, selected variants can
interact and interfere with selection acting for or against other mutations. In
their foundational paper, @Hill1966-gv recognized that linked negatively
selected variants impede the efficacy of a population to remove those
mutations. Generally, linked selection reduces the fixation probability of
advantageous mutations and increases that of deleterious mutations, compared to
expectations under single-locus dynamics [@Birky1988-jm; @Barton1995-mj]. The
joint allele frequency dynamics of linked variants under selection are also
predicted to deviate from expectations without interference. Since haplotypes
that carry multiple deleterious mutations have lower fitness than haplotypes
that carry just one or the other under a multiplicative fitness model, we
should expect to see those mutation segregate on different haplotypes more
often than together, leading to negative, or repulsion, linkage disequilibrium
(LD), although the extent of LD depends non-trivially on the strength of
selection and probability of recombination separating loci. Nonetheless, this
provides a simple, testable hypothesis under the standard selection model of
additive effects within loci and multiplicative fitness between: we should
observe average signed LD to be negative between pairs of selected variants.

Aside from selective interference under a multiplicative fitness model, where
the relative fitness conferred by a haplotype carrying two deleterious
mutations is equal to the product of the fitnesses of haplotypes carrying
individual mutations, considerable attention has been given to
non-multiplicative deviations in the fitness function, known as epistasis.
Early work focused primarily on mutations of strong effect and explored models
of both synergistic and diminishing returns (or antagonistic or threshold)
epistasis under sexual and asexual reproduction, often in the limiting case of
free recombination between deleterious variants. Generally, for sexually
reproducing organisms, deleterious load is decreased under simple models of
synergistic epistasis compared to multiplicative fitness [@Kimura1966-gc;
@Kondrashov1995-va]. Pervasive epistasis has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], as well as driving incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero, with negative synergistic epistasis leading to an excess of negative LD
[@Kondrashov1982-sf; @Charlesworth1990-kw].

While epistasis is the most commonly considered nonlinear effect influencing
interactions between selected variants, non-additive effects *within* a
locus may also cause deviations from our baseline models, either with or
without epistasis. Dominance and multi-locus interactions are jointly important
in shaping the expected equilibrium allele frequencies and mutation load due to
strongly damaging disease mutations [@Clark1998-kq]. However, despite dominance
and partial recessiveness being appreciated as important factors shaping linked
variation [@Turelli2000-kz; @Zhao2016-bb], dominance has received less
attention than epistasis in multi-locus selection models, and its impact on the
joint segregation of negatively selected variants is poorly understood. There
is ample evidence that a large fraction of selected mutations are at least
partially recessive in humans [@xx], flies [@xx], and other species [@xx], and
that average levels of dominance vary with the strength of selection [@xx].

- Patterns of dominance
  - @Huber2018-cp, @Haldane1930-tt, @Kacser1981-nc

### @Ohta1969-ie and @Ohta1971-yd

- Diffusion models for the neutral two-locus decay of LD $\sigma_d^2$

### @Hallgrimsdottir2008-sq

-

### @Good2020-wj

-

- Neutrality, additive, dominance, epistasis, general diploid selection models
- Analytic results: Older stuff, like some basic results from HR, through more
  recent results in @Good2020-wj
  - recursions under neutrality [@Golding1984-pu; @Hudson2001-sg]
- Recent numerical approaches [@Ragsdale2017-gg; @Ragsdale2019-nt] and others that aren't me?
- Resorting to discrete forward simulations (a few examples)

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations within and between protein-coding regions, and
assess how well we can expect to discriminate between modes of selective
interactions from genome-wide patterns of LD.

# Methods

## The two-locus sampling distribution with arbitrary selection

### Drift and recombination

### Selection models with epistasis and dominance

### Low-order summaries of the sampling distribution

## Numerical solution

### Moment closure approach

### Implementation

- Numerical approach for solving the two-locus model with recombination,
  single-population demography, and general selection models, following
  @Jouganous2017-pq and @Ragsdale2019-nt.

### Validation

### Expected LD under inferred human demographic history

## Analysis of human genomic data
- Computing LD statistics commonly used for inference about selection models
  (Figure \@ref(fig:fig1)), using @Ragsdale2020-nz
- Data analysis of Thousand Genomes Project data
  @1000_Genomes_Project_Consortium2015-zq
- Domain information from @Stanek2020-pa

**idea**: LD within genes partitioned by whether two mutations fall within
domains or are in different domains (distance as a proxy? or can we get
information about whether they lie in the same gene or not?)

## Data and software availability

- Thousand genomes VCFs and ancestral sequence
  [@1000_Genomes_Project_Consortium2015-zq]
- Protein domain information [@Stanek2020-pa]
- Implementation as \texttt{moments.TwoLocus} at
  \url{https://bitbucket.org/simongravel/moments}
- Scripts to recreate analyses, figures, and compile this manuscript at
  \url{https://github.com/apragsdale/two_locus_selection}

# Results

## Singed LD under steady-state demography

- Expected LD under models of epistasis and dominance
  at equilibrium (Figures \@ref(fig:fig2) and \@ref(fig:fig3))
  - $\sigma_d^1$, $\sigma_d^2$, $r^2$, for all frequencies
  - The same statistics conditioned on sample allele counts
  - Relationship between "Hudson slice" and allele count-conditioned LD

### Additive selection and epistasis

### Simple dominance within a locus

### General selection and gene-based dominance

## Population size changes [[do what]]

- Effect of demography on these same statistics
  - Bottleneck with recovery
  - Human-like size histories (using @Speidel2019-nj)

## Data... main result in title

- Present data from loss-of-function (LOF), synonymous, and nonsynonymous
  variants from Thousand Genomes, both for pairs of loci within genes and for
  pairs of loci between nearby genes. Do we see any differences, and do the
  different patterns in African vs non-African populations tell us anything
  about the possible selection model for different categories of variants?

# Discussion

- What can we say about epistasis and models of dominance using LD statistics?
  Do we have power to distinguish models using low-order statistics of this
  kind?
- The role of numerical solutions to such models for population genetics
  analyses. They bridge the gap between theory limited by analytic
  intractability and expensive forward simulations such as \texttt{SLiM} and
  \texttt{fwdpy11}.
- Limitations and future directions:
  - multiple populations
  - strong selection leads to numerical instability

# Acknowledgements
- Simon Gravel
- Kevin Thornton

# Figures:

```{r fig1, echo=F, fig.cap="(ref:fig1cap)", fig.align="center"}
knitr::include_graphics("figures/fig1.pdf")
```

(ref:fig1cap) \textbf{Additive selection and allele count-conditioned LD.} (A
and B) The distribution of $AB$ haplotype counts in a sample size of 30, in
which we observe 10 $A$ alleles at the left locus, and 10 $B$ alleles at the
right locus. (C and D) The decay of normalized squared LD ($\sigma_d^2$) with
scaled recombination distance for pairs of neutral and selected variants,
respectively. The dashed line is the neutral expectation [@Ohta1971-yd]. (E and
F) Similarly, the decay of $\sigma_d^1=\mathbb{E}[D]/ \mathbb{E}[p(1-p)q(1-q)]$
for neutral and selected variants, respectively. Note the difference in scale
between the two panels.

```{r fig2, echo=F, fig.cap="(ref:fig2cap)", fig.align="center"}
knitr::include_graphics("figures/fig2.pdf")
```

(ref:fig2cap) \textbf{Additive selection with epistasis.} (A) $D^2$, and (B) $D$.

```{r fig3, echo=F, fig.cap="(ref:fig3cap)", fig.align="center"}
knitr::include_graphics("figures/fig3.pdf")
```

(ref:fig3cap) \textbf{Dominance.} (A) $D^2$, and (B) $D$.

4. Effects of demography on LD (bottleneck and recovery vs expansion, following
   Tennessen model)

5. LOF, missense, and synonymous variants, within and between genes.

6. Missense variants in genes partitioned by conservation? Tolerance to
   nonsynonymous changes?
